---
title: 'Predicting Wages: Polynomial Regression & Step Functions'
author: "By Vishwa Pardeshi"
output:
  pdf_document: default
  html_document: default
---

In this notebook, we will predict the wages of males who reside in the central Atlantic region of the United States. 

**Learning Outcome:**
By following the notebook you will be able to 

1. Implement Polynomial Regression & Step Funtion

2. Identify optimal degree using cross-validation

3. Perform hypothesis testing using ANOVA

## Setup
```{r setup}
library(ISLR)
library(boot)
```

## Find optimal degree for polynomial regression.

```{r cross validation}
set.seed(1)

cv.error <- rep(0,10)
for (i in 1:10) {
  glm.fit <- glm(wage~poly(age,i), data=Wage)
  cv.error[i] <- cv.glm(Wage, glm.fit, K=10)$delta[1]  # [1]:std, [2]:bias-corrected
}
cv.error
```

### Plot the cross validation error

```{r pressure, echo=FALSE}
plot(cv.error, type="b")
```

The optimal degree for polynomial regression model is 9 as it lowest cross validation error.

## Hypothesis Testing using ANOVA

```{r ANOVA}
fit.01 <- lm(wage~age, data=Wage)
fit.02 <- lm(wage~poly(age,2), data=Wage)
fit.03 <- lm(wage~poly(age,3), data=Wage)
fit.04 <- lm(wage~poly(age,4), data=Wage)
fit.05 <- lm(wage~poly(age,5), data=Wage)
fit.06 <- lm(wage~poly(age,6), data=Wage)
fit.07 <- lm(wage~poly(age,7), data=Wage)
fit.08 <- lm(wage~poly(age,8), data=Wage)
fit.09 <- lm(wage~poly(age,9), data=Wage)
fit.10 <- lm(wage~poly(age,10), data=Wage)
anova(fit.01,fit.02,fit.03,fit.04,fit.05,fit.06,fit.07,fit.08,fit.09,fit.10)
```

ANOVA hypothesis shows statistically significant result for degree 3 and 9. Since we noticed that the cv error for degree 9 was the lowest, we pick degree 9 as the optimal degree for our polynomial regression model.

```{r}
agelims <- range(Wage$age)
age.grid <- seq(agelims[1], agelims[2])
preds <- predict(fit.09, newdata=list(age=age.grid), se=TRUE)
se.bands <- preds$fit + cbind(2*preds$se.fit, -2*preds$se.fit)
par(mfrow=c(1,1), mar=c(4.5,4.5,1,1), oma=c(0,0,4,0))
plot(Wage$age, Wage$wage, xlim=agelims, cex=0.5, col="darkgrey")
title("Degree 9 Polynomial Fit", outer=TRUE)
lines(age.grid, preds$fit, lwd=2, col="blue")
matlines(age.grid, se.bands, lwd=1, col="blue", lty=3)
```

## Find optimal cut for step function using CV

```{r optimal cut }
set.seed(1)

# cross-validation
cv.error <- rep(0,9)
for (i in 2:10) {
  Wage$age.cut <- cut(Wage$age,i)
  glm.fit <- glm(wage~age.cut, data=Wage)
  cv.error[i-1] <- cv.glm(Wage, glm.fit, K=10)$delta[1]  # [1]:std, [2]:bias-corrected
}
cv.error
```

### Plot for cv error
```{r}
plot(2:10, cv.error, type="b")
```

## Implement step functions

```{r step function}
cut.fit <- glm(wage~cut(age,8), data=Wage)
preds <- predict(cut.fit, newdata=list(age=age.grid), se=TRUE)
se.bands <- preds$fit + cbind(2*preds$se.fit, -2*preds$se.fit)
plot(Wage$age, Wage$wage, xlim=agelims, cex=0.5, col="darkgrey")
title("Fit with 8 Age Bands")
lines(age.grid, preds$fit, lwd=2, col="blue")
matlines(age.grid, se.bands, lwd=1, col="blue", lty=3)
```
